---
title: Choosing a new default RNG for dqrng
author: ''
date: '2023-09-09'
slug: choosing-a-new-default-rng-for-dqrng
categories:
  - R
tags:
  - R
  - RNG
  - package
image:
  caption: ''
  focal_point: ''
bibliography: references.bib
---

```{r, echo=FALSE, include=FALSE}
library(bench)
library(tidyr)
```


Currently the dqrng package supports only xoroshiro128+ and xoshiro256+ from <https://prng.di.unimi.it/> [see also @blackman2021]. These RNGs should only be used for creating floating point numbers, which was the case for dqrng originally. However, `dqsample` and `dqrrademacher` make use of the full bit pattern. So it would be better to support the `**` and/or `++` variants for both RNGs and make one of them the default. This would be a breaking change, of course. In [#57](https://github.com/daqana/dqrng/pull/57) I have added these additional 4 RNGs to `xoshiro.h` so now is the time to do some benchmarking first by generating some random numbers:

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <dqrng_distribution.h>
#include <xoshiro.h>
// [[Rcpp::plugins(cpp11)]]

template<typename RNG>
double sum_rng(int n) {
  auto rng = dqrng::generator<RNG>(42);
  dqrng::uniform_distribution dist;
  double result = 0.0;
  for (int i = 0; i < n; ++i) {
    result += dist(*rng);
  }
  return result;
}

// [[Rcpp::export]]
double sum_128plus(int n) {
  return sum_rng<dqrng::xoroshiro128plus>(n);
}
// [[Rcpp::export]]
double sum_256plus(int n) {
  return sum_rng<dqrng::xoshiro256plus>(n);
}
// [[Rcpp::export]]
double sum_128starstar(int n) {
  return sum_rng<dqrng::xoroshiro128starstar>(n);
}
// [[Rcpp::export]]
double sum_256starstar(int n) {
  return sum_rng<dqrng::xoshiro256starstar>(n);
}
// [[Rcpp::export]]
double sum_128plusplus(int n) {
  return sum_rng<dqrng::xoroshiro128plusplus>(n);
}
// [[Rcpp::export]]
double sum_256plusplus(int n) {
  return sum_rng<dqrng::xoshiro256plusplus>(n);
}
```

```{r}
N <- 1e5
bm <- bench::mark(
  sum_128plus(N),
  sum_128starstar(N),
  sum_128plusplus(N),
  sum_256plus(N),
  sum_256starstar(N),
  sum_256plusplus(N),
  check = FALSE,
  min_time = 1
)
knitr::kable(bm[, 1:6])
plot(bm)
```
The current default xoroshiro128+ is fastest in this comparison with the other generators being very similar. Let's try a more realistic usecase: generating many uniformaly distributed random numbers:

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <dqrng_distribution.h>
#include <xoshiro.h>
// [[Rcpp::plugins(cpp11)]]

template<typename RNG>
Rcpp::NumericVector runif_rng(int n) {
  auto rng = dqrng::generator<RNG>(42);
  dqrng::uniform_distribution dist;
  Rcpp::NumericVector result(Rcpp::no_init(n));
  std::generate(result.begin(), result.end(), [rng, dist] () {return dist(*rng);});
  return result;
}

// [[Rcpp::export]]
Rcpp::NumericVector runif_128plus(int n) {
  return runif_rng<dqrng::xoroshiro128plus>(n);
}
// [[Rcpp::export]]
Rcpp::NumericVector runif_256plus(int n) {
  return runif_rng<dqrng::xoshiro256plus>(n);
}
// [[Rcpp::export]]
Rcpp::NumericVector runif_128starstar(int n) {
  return runif_rng<dqrng::xoroshiro128starstar>(n);
}
// [[Rcpp::export]]
Rcpp::NumericVector runif_256starstar(int n) {
  return runif_rng<dqrng::xoshiro256starstar>(n);
}
// [[Rcpp::export]]
Rcpp::NumericVector runif_128plusplus(int n) {
  return runif_rng<dqrng::xoroshiro128plusplus>(n);
}
// [[Rcpp::export]]
Rcpp::NumericVector runif_256plusplus(int n) {
  return runif_rng<dqrng::xoshiro256plusplus>(n);
}
```

```{r}
N <- 1e5
bm <- bench::mark(
  runif(N),
  runif_128plus(N),
  runif_128starstar(N),
  runif_128plusplus(N),
  runif_256plus(N),
  runif_256starstar(N),
  runif_256plusplus(N),
  check = FALSE,
  min_time = 1
)
knitr::kable(bm[, 1:6])
plot(bm)
```
Here all six generators are very similar, with all of them clearly faster than R's built in `runif`. How about sampling with replacement, which is also mostly governed by the speed of generating random numbers:

```{Rcpp}
#include <Rcpp.h>
// [[Rcpp::depends(dqrng, BH)]]
#include <dqrng_sample.h>
#include <xoshiro.h>
// [[Rcpp::plugins(cpp11)]]

// [[Rcpp::export]]
Rcpp::IntegerVector sample_128plus(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoroshiro128plus>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
// [[Rcpp::export]]
Rcpp::IntegerVector sample_128starstar(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoroshiro128starstar>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
// [[Rcpp::export]]
Rcpp::IntegerVector sample_128plusplus(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoroshiro128plusplus>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
// [[Rcpp::export]]
Rcpp::IntegerVector sample_256plus(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoshiro256plus>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
// [[Rcpp::export]]
Rcpp::IntegerVector sample_256starstar(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoshiro256starstar>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
// [[Rcpp::export]]
Rcpp::IntegerVector sample_256plusplus(int m, int n) {
  auto rng = dqrng::generator<dqrng::xoshiro256plusplus>(42);
  return dqrng::sample::sample<INTSXP, uint32_t>(rng, uint32_t(m), uint32_t(n), true, 0);
}
```

```{r}
N <- 1e5
M <- 1e3
bm <- bench::mark(
  sample.int(M, N, replace = TRUE),
  sample_128plus(M, N),
  sample_128starstar(M, N),
  sample_128plusplus(M, N),
  sample_256plus(M, N),
  sample_256starstar(M, N),
  sample_256plusplus(M, N),
  check = FALSE,
  min_time = 1
)
knitr::kable(bm[, 1:6])
plot(bm)
```
Again nothing really conclusive. All six RNGs are similar and much faster than R's build in `sample.int`.

The speed comparisons between these generators is inconclusive to me. The xoroshiro128 seem to be slightly faster than the xoshiro256 variants. So I am leaning towards one of those as the new default while still making the corresponding xoshiro256 variant available in case a longer period or a higher degree of parallelisation is needed. Comparing the `++` and `**` variants, I am leaning towards `++`, but that is not set in stone.

## References
